Experience Report: Haskell in Computational Biology

### 2012-06-27 mrfy_experience_report.tex
edit the intro (first column) for verbs and for parallel structure

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language along with performance comparable to~C++.
In~our experience, higher-order functions, lazy evaluation, and
monads really worked, but
profiling and debugging presented obstacles.
Haskell libraries varied greatly:
libraries providing memoization combinators and parallel-evaluation
strategies gave us tremendous leverage,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a high-level
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although we had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and parallel performance was easy to achieve
(\secref{perf}). 
\item
Higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
The only significant impediment that Haskell presented to our
research was the difficulty of understanding and improving the
performance of Haskell programs (\secref{awkward-profiling}).
\item
We~used Haskell successfully even though we are computational
biologists with little background in functional programming.
Modifying our group's C++ code to try new research ideas has been
difficult;
trying new ideas in Haskell has been easy
(\secref{comparo}).
\item
The Haskell language is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time.
We~found the two families difficult to distinguish in advance (\secref{penumbra}).
\end{itemize}

>>>>

%
% It can't hurt to remind everyone involved exactly what an Experience
% Report is.
%
%
\begin{abstract}
%An Experience Report provides {evidence} that functional
%programming really works, or it describes obstacles that prevent
%functional programming from working.   
Haskell gives computational biologists
the flexibility and rapid prototyping of a scripting
language along with performance comparable to~C++.
In~our experience, higher-order functions, lazy evaluation, and
monads really worked, but
profiling and debugging presented obstacles.
Haskell libraries varied greatly:
libraries providing memoization combinators and parallel-evaluation
strategies helped us a lot,
but other, nameless libraries mostly got in our way.
Despite the obstacles and the uncertain quality of some libraries,
Haskell's ecosystem
made it easy for us to develop new algorithms in computational
biology.
\end{abstract}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
but the implementation required little effort,
and the connection between code and
algorithm is unusually clear (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and it was easy to parallelize
(\secref{perf}). 
\remark{Can we summarize the performance results?  Speedup?}
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one way: it's hard to understand and
improve the
performance of a Haskell program (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although authors Daniels and Gallant are computational
biologists with little background in functional programming,
we~used Haskell successfully.
And Haskell has made it easy to explore new research ideas;
by~contrast,
our group's C++ code has not been a good vehicle for exploring new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some worked as advertised, saved us lots of time, and were a pleasure
to~use.
Others, not so much.
And~we could not tell ahead of time which experience we were likely to have
 (\secref{penumbra}).
\end{itemize}
